Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-03-30 doc/thesis/objective.tex
Small changes to objective

\itshape
This section is intended to be a synthesis of all the background material
together into some research objectives.
\normalfont

The spread of a disease through a susceptible population is most often modelled
by assuming either a homogeneously mixed population, or a population divided
into a small number of homogeneously mixed compartments. This assumption, also
called the ``law of mass action'', implies that any two individuals in the same
compartment are equally likely to come into contact causing transmission.
Although this is clearly unrealistic for most populations, the predictions they
make have proven to be fairly accurate in practice. However, when there is
substantial contact heterogeneity, such as when considering a sexually
transmitted infection, the error introduced by assuming a panmictic population
may be substantial.

Contact network models provide an alternative to compartmental models which do
not require the assumption of panmixia. In addition to more accurate
prevalence predictions, the parameters of the networks themselves may be of
interest from a public health perspective. For example, certain vaccination
strategies may be more or less effective in curtailing an epidemic depending on
the underlying network's degree distribution.

Phylodynamic methods have been used to investigate a great diversity of
epidemiological parameters. However, as far as we are aware, no phylodynamic
methods have been developed to fit contact network models. The primary
objective of this work is to develop such a method.

Network structure has a substantial impact on the speed and pattern of epidemic
spread. Indeed, a small number of methods have been developed which exploit
this fact to perform statistical inference. Contact network parameters, such as
edge density and degree distribution, have been estimated from epidemiological
data, such as incidence reports and individual recovery times.

Contact networks' parameters can have a strong effect on the shape of
transmission trees, and therefore on viral phylogenies, resulting from
epidemics on those networks. Yet, this relationship has not yet been
``inverted'' to quantitatively estimate network parameters from transmission
trees.

@@

Although we have not proven it analytically, calculating the likelihood of the
parameters of a contact network models seems likely to be an intractable
problem. This would imply that these models are amenable to neither \gls{ML}
nor Bayesian inference. To illustrate this, consider a contact network model
with parameters $\theta$, and an observed transmission tree $T$ with $n$ tips.
In general, we do not know the labels of the internal nodes of $T$, only the
labels of its tips. To fit this model using likelihood-based methods, we must
calculate the likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$
be the set of all possible contact networks, and $\Nu$ be the set of all
possible labellings of the internal nodes of $T$. We can write the likelihood as
\begin{align}
\begin{split}
  \label{eq:netlik}
  \Pr(T \mid \theta)
    &= \sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),
\end{split}
\end{align}
the last equality following from the fact that $T$ and $\nu$ depend only on
$G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
may individually be straightforward to calculate, the number of possible
directed graphs on $N$ nodes is $2^{N(N-1)}$, larger if the nodes and edges in
the graph may have different labels or attributes. Hence, the number of terms
in the sum is at least exponential in $n$, more if we do not know \textit{a
priori} how many nodes are in the network (as is likely). In addition,
\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
individuals were sampled. This is rarely the case in practice - most often, the
observed tree is a subsampled version of the true tree. In this case, the
likelihood calculation becomes even more complex, because we must also sum over
all possible complete trees.

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative 
to likelihood-based methods, which would apply to any network model, is
provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} are
readily available. Simulating networks is straightforward under a variety of
models. Epidemics on those networks, and the corresponding transmission trees,
can also be easily simulated. As mentioned above, contact networks can
profoundly affect transmission tree shape, and those shapes can be compared
using a highly informative similarity measure. \Gls{SMC} has several advantages
over other algorithms for \gls{ABC}~\autocite{mckinley2009inference}, including
a recently-developed adaptive algorithm requiring minimal tuning on the part of
the user~\autocite{del2012adaptive}. In summary, our method to infer contact
network parameters will combine the following: stochastic simulation of
epidemics on networks, the tree kernel, and adaptive \gls{ABC}-\gls{SMC}.

>>>>

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population, or a population divided into a small number of
homogeneously mixed compartments. This assumption, also called the ``law of
mass action'', implies that any two individuals in the same compartment are
equally likely to come into contact causing transmission. Although this
provides a reasonable approximation in many cases, the error introduced by
assuming a panmictic population can be substantial when significant contact
heterogeneity exists in the underlying population. Contact network models
provide an alternative to compartmental models which do not require the
assumption of panmixia. In addition to more accurate predictions, the
parameters of the networks themselves may be of interest from a public health
perspective. For example, certain vaccination strategies may be more or less
effective in curtailing an epidemic depending on the underlying network's
degree distribution. Phylodynamic methods have been used to fit many different
types of model to phylogenetic data, but as far as we know, no methods have yet
been developed to fit contact network models. The primary objective of this
work is to develop such a method.

@@

Calculating the likelihood of the parameters of a contact network models seems
likely to be an intractable problem, which would imply that these models are
amenable to neither \gls{ML} nor Bayesian inference. We not proven this is the
case, but some intuition can be provided by examining the process involved in
the likelihood calculation. Consider a contact network model with parameters
$\theta$, and an observed transmission tree $T$ with $n$ tips.  In general, we
do not know the labels of the internal nodes of $T$, only the labels of its
tips. To fit this model using likelihood-based methods, we must calculate the
likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of
all possible contact networks, and $\Nu$ be the set of all possible labellings
of the internal nodes of $T$. We can write the likelihood as
\begin{align}
\begin{split}
  \label{eq:netlik}
  \Pr(T \mid \theta)
    &= \sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),
\end{split}
\end{align}
the last equality following from the fact that $T$ and $\nu$ depend only on
$G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
may individually be straightforward to calculate, the number of possible
directed graphs on $N$ nodes is $2^{N(N-1)}$, larger if the nodes and edges in
the graph may have different labels or attributes. Hence, the number of terms
in the sum is at least exponential in $n$, more if we do not know \textit{a
priori} how many nodes are in the network (as is likely). In addition,
\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
individuals were sampled. This is rarely the case in practice - most often, the
observed tree is a subsampled version of the true tree. In this case, the
likelihood calculation becomes even more complex, because we must also sum over
all possible complete trees.

Depending on the network model studied, it is possible that \cref{eq:netlik}
could be simplified into a tractable expression. However, a simpler alternative
to likelihood-based methods, which would apply to any network model, is
provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} to
this problem are readily available. Simulating networks is straightforward
under a variety of models. Epidemics on those networks, and the corresponding
transmission trees, can also be easily simulated. As mentioned above, contact
networks can profoundly affect transmission tree shape, and those shapes can be
compared using a highly informative similarity measure. \Gls{SMC} has several
advantages over other algorithms for \gls{ABC}~\autocite{mckinley2009inference},
including a recently-developed adaptive algorithm requiring minimal tuning on
the part of the user~\autocite{del2012adaptive}. In summary, our method to
infer contact network parameters will combine the following: stochastic
simulation of epidemics on networks, the tree kernel, and adaptive
\gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
method is a type of kernel-\gls{ABC}. For ease of exposition, we will often use
the term ``kernel-\gls{ABC}'' to refer to our method specifically.

Empirical studies of sexual contact networks have found that these networks
tend to be scale-free, meaning that their degree distributions follow a power
law (although there has been some disagreement). Preferential attachment has
been postulated as a mechanism by which scale-free networks could be generated.
This makes the \gls{BA} model, one of the simplest preferential attachment
models, a natural choice to explore with our method. The second aim of this
work is to use simulations to investigate the parameters of the \gls{BA} model,
including whether they have a detectable impact on tree shape, and whether they
can be accurately recovered using kernel-\gls{ABC}.

Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
the most commonly-studied viruses in a phylodynamic context. Consequently, a
large volume of \gls{HIV} sequence data is publicly available, more than for
any other pathogen, and including sequences sampled from diverse geographic and
demographic contexts. Since \gls{HIV} is almost always spread through either
sexual contact or sharing of injection drug supplies, the contact networks
underlying \gls{HIV} epidemics are highly structured. Moreover, since no cure
yet exists, efforts to curtail the progression of an epidemic have relied on
preventing further transmissions through measures such as \gls{tasp} and
education leading to behaviour change. The effectiveness of this type of
intervention can vary significantly based on the underlying structure of the
network and the particular nodes to whom the intervention is targeted. Due to
this combination of data availability and potential public health impact,
\gls{HIV} is an obvious context in which our method could be applied.
Therefore, the third and final aim of this work is to apply kernel-\gls{ABC} to
fit the \gls{BA} model to existing \gls{HIV} outbreaks.

To summarize, this work has three objectives. First, we will develop a method
which uses kernel-\gls{ABC} to infer parameters of contact network models from
observed transmission trees. Second, we will use simulations to characterize
the parameters of the \gls{BA} network model in terms of their effect on tree
shape and how accurately they can be recovered with kernel-\gls{ABC}. Finally,
we will apply the method fit the \gls{BA} model to several real-world \gls{HIV}
datasets.

### 2016-03-30 doc/thesis/results.Rnw
Small changes to objective

<<treestats, include=FALSE>>=
    params <- c("alpha", "I", "m", "N")
    ptn <- paste0("../../simulations/kernel-", params, "/statistics/*")
    files <- lapply(ptn, Sys.glob)
    d <- lapply(files, collect.data)
    d <- mapply(cbind, d, parameter=as.list(params), SIMPLIFY=FALSE)
    lapply(d, setDT)
    d[[1]] <- subset(d[[1]], m == 2)
    setnames(d[[1]], "nsimnode", "I")
@

>>>>

<<treestats, include=FALSE>>=
    #params <- c("alpha", "I", "m", "N")
    #ptn <- paste0("../../simulations/kernel-", params, "/statistics/*")
    #files <- lapply(ptn, Sys.glob)
    #d <- lapply(files, collect.data)
    #d <- mapply(cbind, d, parameter=as.list(params), SIMPLIFY=FALSE)
    #lapply(d, setDT)
    #d[[1]] <- subset(d[[1]], m == 2)
    #setnames(d[[1]], "nsimnode", "I")
@
